---
title: "Ramen ratings analysis"
output: html_notebook
---
```{r}
ramen_ratings <- read.csv("ramen-ratings.csv")
```

# About the dataset

The chosen dataset exposes the ratings of different types of Ramen from various brands in multiple countries. More specifically, it provides the following relevant information:

- Brand : from which brand that specific ramen is from
- Style : the type of ramen / its serving form (cup, pack, bowl, ...)
- Country : the country where a certain rating came from
- Stars : rating given to the ramen on a five point scale

There are a few more columns that require different handling

- Review number : index of a review which does not provide relevant information for our statistical analysis
- Variety : string containing a description of the type (contents) of ramen, which requires additional processing such as keyword extraction to be practical in our analysis
- Top ten : the year a certain ramen made it to the top ten, along with its position in the ranking

## Initial analysis

To make the data-set more approachable, I conducted some visualisations to better understand the type and layout of the data I am working with. For this however, I first needed to convert the ratings into numbers (float types), as by default they are of string type. Then I proceeded to remove rows with a missing rating to prevent erroneous data in future manipulations/calculations.

```{r}
library(dplyr)
processed_data <- ramen_ratings[, c(2,4,5,6)]
processed_data$Stars <- as.numeric(processed_data$Stars)
processed_data <- na.omit(processed_data)
```

### Visualising the number of ratings

```{r}
rating_freq <- table(processed_data$Stars)
barplot(rating_freq, las=2, main="Frequencies of discrete ratings", ylab="Count", cex.names=.7)
```

By taking brackets: [0,1[, [1, 2[, ...

```{r}
rating_freq_ranges <- table(floor(processed_data$Stars))
barplot(rating_freq_ranges, names=c("0-0.9", "1-1.9", "2-2.9", "3-3.9", "4-4.9", "5"), main="Frequencies of ratings by brackets", ylab="Count")
```


```{r}
options(dplyr.summarise.inform = FALSE)
country_freq <- group_by(processed_data, Country) %>% summarise(Count=n()) %>% arrange(desc(Count))
barplot(country_freq$Count, names=country_freq$Country, las=2, main="Number of ratings per country", ylab="Count", cex.names=.8)
```
